Where Do Events Come From?

### 2018-08-15 paper.tex
(>) Revisões Sztajnberg 2.2 e Conclusão

In \CEU, when multiple trails awake in the same reaction, they execute
atomically in lexical order, i.e., in the order they appear in the source code.
%
In Figure~\ref{lst.mem}, both listings define a shared variable (ln~3) and
assign to it in parallel trails (ln~6, 9).
%
In Listing~\ref{lst.mem.a}, the two assignments to \code{x} can only execute in
reactions to different events \code{A} and \code{B} (ln~5,8), which cannot
occur simultaneously due to the atomicity property of the synchronous model.
%Hence, for the sequence of events \code{A->B}, \code{x} becomes
%\code{4}~(\code{(1+1)*2}), while for \code{B->A}, \code{x} becomes
%\code{3}~(\code{(1*2)+1}).
In Listing~\ref{lst.mem.b}, the two assignments to \code{y} are simultaneous
because they execute in reaction to the same event \code{A}.
Nevertheless, because \CEU follows lexical order and executes atomically, the
outcome is still deterministic, and \code{y} always becomes \code{4}
(\code{(1+1)*2}).
%
Even so, \CEU performs (optional) concurrency checks at compile time to detect
conflicting accesses to shared variables:
if a variable is written in a trail segment, then a concurrent trail segment
cannot access that variable~\cite{ceu.sensys13}.
%
%The static checks are optional (or raise warnings) and do not affect the
%semantics of the language.

The addition of asynchronous ISRs now poses real threats concerning race conditions
since they interrupt programs at arbitrary points.
%
Listing~\ref{lst.usart} illustrates a minimum USART application (ln~13--22)
that consumes incoming bytes (ln~18--21) as they arrive (ln~16).
%
The USART API (ln~1--4) exposes an input event to signal incoming data (ln~3)
and a buffer to prevent data loss (ln~4).
The buffer is indispensable because the microcontroller might not cope with the USART 
speed.
%
The driver ISR (ln~6--11) simply appends incoming bytes to the end of the
buffer (ln~9) and signals the application (ln~10).
%
As a possible race condition, note that the ISR may fire and append a new byte
to the buffer (ln~9) just before the application clears it (ln~21), in which
case the new byte will be lost forever.

@@

@@

loop do
    await USART_RX;
    var int i;
    loop i in [0 -> $rx_buf[ do   // `$' = length ($)
        // uses rx_buf[i]
    end
    $rx_buf = 0;
end
\end{lstlisting}

@@

Note that applications never need to be explicit about energy management to take
advantage of sleep modes.
All happens automatically because of the synchronous reactive execution model
and the interaction between the drivers and energy-aware runtime of \CEU.

@@

Our approach for developing drivers relies on the lexical structured mechanisms of
\CEU, such as parallel compositions and abortion of lines of execution, to
offer stronger guarantees.
%
For instance, the visibility of drivers can be delimited to scoped blocks to
avoid resource leaks.
%
The same mechanism allows drivers to signal the power manager that associated
peripherals are no longer required, allowing applications to use optimal
sleep modes automatically.
%
Our initial tests show optimal energy savings for applications in idle states.

>>>>

In \CEU, when multiple trails awake in the same reaction, they execute
atomically in lexical order, i.e., in the order they appear in the source code.
%
In Figure~\ref{lst.mem}, both listings define a shared variable (ln~3) which
is assigned in parallel trails (ln~6, 9).
%
In Listing~\ref{lst.mem.a}, the two assignments to \code{x} can only execute in
reactions to different events \code{A} and \code{B} (ln~5,8), which cannot
occur simultaneously due to the atomicity property of the synchronous model.
%Hence, for the sequence of events \code{A->B}, \code{x} becomes
%\code{4}~(\code{(1+1)*2}), while for \code{B->A}, \code{x} becomes
%\code{3}~(\code{(1*2)+1}).
In Listing~\ref{lst.mem.b}, the two assignments to \code{y} are simultaneous
because they execute in reaction to the same event \code{A}.
Nevertheless, because \CEU follows lexical order and executes atomically, the
outcome is still deterministic, and \code{y} always becomes \code{4}
(\code{(1+1)*2}).
%
Even so, \CEU performs (optional) concurrency checks at compile time to detect
conflicting accesses to shared variables:
if a variable is written in a trail segment, then a concurrent trail segment
cannot access that variable~\cite{ceu.sensys13}.
%
%The static checks are optional (or raise warnings) and do not affect the
%semantics of the language.

The addition of asynchronous ISRs now poses real threats concerning race
conditions since they interrupt programs at arbitrary points and could possibly
share variables with synchronous code.
%
Listing~\ref{lst.usart} illustrates a minimum USART application (ln~13--22)
that consumes incoming bytes (ln~18--21) as they arrive (ln~16).
%
The USART API (ln~1--4) exposes an input event to signal incoming data (ln~3)
and a buffer to prevent data loss (ln~4).
The buffer is indispensable because the microcontroller might not cope with the USART 
speed.
%
The driver ISR (ln~6--11) simply appends incoming bytes to the end of the
buffer (ln~9) and signals the application (ln~10).
%
As a possible race condition, note that the ISR may fire and append a new byte
to the buffer (ln~9) just before the application clears it (ln~21), in which
case the new byte will be lost forever.

@@

spawn async/isr [USART_RX_vect] do
    rx_buf = rx_buf .. [{UDR0}];  // appends new byte
    if $rx_buf == 1 then
        emit USART_RX; // emits unless already pending
    end                // (assumes client consumes all)
end

@@

loop do
    await USART_RX;
    var int i;
    loop i in [0 -> $rx_buf[ do   // `$' = length
        // uses rx_buf[i]
    end
    $rx_buf = 0;
end
\end{lstlisting}

@@

Note that applications never need to be explicit about energy management to take
advantage of sleep modes.
All happens automatically because of the synchronous reactive execution model
and the interaction between the drivers and energy-aware runtime of \CEU.
%
Our initial tests show optimal energy savings for applications in idle
states~\cite{ceu.lctes18.short}.

@@

Our approach for developing drivers relies on the lexical structured mechanisms of
\CEU, such as parallel compositions and abortion of lines of execution, to
offer stronger guarantees.
%
For instance, the visibility of drivers can be delimited to scoped blocks to
avoid resource leaks.
%
The same mechanism allows drivers to signal the power manager that associated
peripherals are no longer required, allowing applications to use optimal
sleep modes automatically.
